Software Product Line Analysis Techniques

### 2018-01-20 doc/quals/quals.tex
Intro

\section{Introduction}
TODO

@@

An FTS extends a Labeled Transition System (LTS) with a total function $\gamma$ mapping state transitions to presence conditions. This way behavior due to different model variants can be guarded by the presence conditions of those variants. In addition, an FTS is nonmonotonic in the sense that existing behavior can be removed by a variant by assigning the negation of the presence condition of that variant to the transitions associated with the behavior.

Parallel composition of FTSs is also defined similarly to that of LTSs. When two FTSs are parallely composed, independent transition actions are asynchronously interleaved. However, shared actions are synchronized and their corresponding presence conditions are conjoined.

@@

Given a programming language $\lang$, starting with an abstract intepretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$. If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.

\emph{Variational abstraction} pushes abstraction up to the configuration space. Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation. Pretty much like all abstraction techniques, this trades accuracy for performance. \emph{Variational abstraction} is implemented the same as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs. Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where collapses the configuration space into a single value). Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.

>>>>

\section{Introduction}
A Software Product Line (SPL) is a family of related software products built from a common set of artifacts (e.g., design/specification models, source code, test cases). An SPL is composed of a set of features. A feature is an externally visible property, aspect or quality of a software system. An instance (product) of an SPL is composed of a subset of the SPL features. Not all combinations of features are valid though. A Feature Model (FM) - also known as a Variability Model (VM) - is a specification of the valid combinations of SPL features. An FM can be expressed as a boolean formula over features, or graphically as a Feature Diagram (FD).

When building an SPL, software analysis tools are still needed to find bugs, generate test cases, collect metrics or track progress. A plethora of tools already exists, covering many different categories of analyses. However, these tools work on the artifacts of a single product. Because an SPL encompasses a superset of features existing in a single product, those tools are not capable of identifying which artifacts belong to which set of products. Consequently, those tools cannot be used as is on an SPL. A naive solution is to generate all possible products of an SPL and run tools on each one of them. This is generally intractable though, because the number of products grows exponentially with the number of SPL features.

This position paper covers some research work attempting to solve this problem. One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques\cite{Apel:2013}. Another extends this work to more elaborate sampling techniques and more general categories of analysis\cite{Liebig:2013}. At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs\cite{Midtgaard:2015}. At the other end, some papers provide pragmatic implementations of source code syntax checkers\cite{Gazzillo:2012}, type checkers\cite{Kastner:2012}, data flow analyses\cite{Bodden:2013}, and model checkers\cite{Classen:2013}.

Given a programming language $\lang$, starting with an abstract interpretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$. If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.

\emph{Variational abstraction} pushes abstraction up to the configuration space. Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation. Pretty much like all abstraction techniques, this trades accuracy for performance. \emph{Variational abstraction} is implemented the same way as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs. Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where the configuration space is collapsed into a single value). Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.

@@

An FTS extends a Labeled Transition System (LTS) with a total function $\gamma$ mapping state transitions to presence conditions. This way behavior due to different model variants can be guarded by the presence conditions of those variants. In addition, an FTS is non-monotonic in the sense that existing behavior can be removed by a variant by assigning the negation of the presence condition of that variant to the transitions associated with the behavior.

Parallel composition of FTSs is also defined similarly to that of LTSs. When two FTSs are composed in parallel, independent transition actions are asynchronously interleaved. However, shared actions are synchronized and their corresponding presence conditions are conjoined.

@@
